---
title: SDS - HW3 - Connect your brain
author: Marco Aurelio Sterpa, Stefania Sferragatta
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r utilities, message=FALSE, warning=FALSE}
library(DescTools, verbose = F)
library(igraph, verbose = F)
library(matrixStats, verbose = F)
load("hw3_data.RData")
```

## Build the graphs & Intersection test
In order to compute the evaluation of the dependency between cortical regions, the starting point is to build an association graph. The vertices are the features of each sample (that are as many as the subjects) and the edges are the relationships between them. 

In this function we build our estimates related to the true association graphs. First of all we define adjustment of nominal level of the intervals with the Bonferroni correction, computed by replacing the $\alpha$ with $\frac{\alpha}{m}$ where m is the binomial coefficient $D \choose 2$ that indicates the number of intervals we are going to build. 

Then for each group we define a threshold given by the $80^{th}$ percentile of the correlation values, it will be used in the test for empty intersection define after this function. 

We start by computing the correlation matrices for each subject and the pool this data together taking the average as they are supposed to be independent. We now have two correlation matrices representing the observed correlation in each group to which we apply the Fisher Z-transform to obtain confidence interval. 

The correlation graphs estimates are built by placing an edge between features with empty intersection between its CI and the interval $[-t,t]$, where $t=q_.8$.
```{r loading, warning=FALSE}
average_cor = function(M) FisherZInv(colMeans(FisherZ(M)))
to_matrix = function(M_flat) matrix(M_flat, nrow=116, byrow=T)

get_adjancency_matrix = function(lower, upper, threshold) {
  M = to_matrix(sapply(1:length(lower), function(i) int.test(lower[i], upper[i], threshold)))
  M[is.na(M)] = 0
  return(M)
}

# test for empty intersection of [-t,t] with CI(j,k)
int.test <- function(lower, upper, t) (t < lower) | (-t > upper)


build_graph = function(A_flat, B_flat, probs = 0.8, adjust = TRUE) {
  # Bonferroni adjusted confidence level
  d = 116
  m = d*(d-1)/2
  if(adjust) alpha = 1 - 0.05 / m
  else alpha = 1 - 0.05
  
  # average correlation per group
  A_means = average_cor(A_flat)
  B_means = average_cor(B_flat)
  
  # thresholds
  t_a = quantile(A_means, probs = probs, names = F, na.rm = T)
  t_b = quantile(B_means, probs = probs, names = F, na.rm = T)
  
  # Fisher confidence intervals
  A_fisher = sapply(A_means, function(x) CorCI(x, n = 145, conf.level = alpha))
  B_fisher = sapply(B_means, function(x) CorCI(x, n = 145, conf.level = alpha))
  
  # adjacency matrices
  n = dim(A_fisher)[2]
  A_matrix = get_adjancency_matrix(lower = A_fisher[2,], upper = A_fisher[3,],
                                   threshold = t_a)
  B_matrix = get_adjancency_matrix(lower = B_fisher[2,], upper = B_fisher[3,],
                                   threshold = t_b)
  
  return(list(A = A_matrix, B = B_matrix))
}
```


Below we define some useful functions used to visualize the estimated graphs. We compare the graph with and without the Bonferroni correction and also with the community detection in order to highlight some hidden relations among the features.  
```{r, warning=FALSE}
set_graph_params = function(G) {
  V(G)$size <- .5
  V(G)$label <- ""
  E(G)$width = 0.01
  E(G)$arrow.mode <- 0
  return(G)
}

plot_graph = function(M, title, community=F) {
  G = simplify(graph_from_adjacency_matrix(M, mode="undirected"), remove.loops = T)
  G = set_graph_params(G)
  params = list(G, main = title, vertex.color="indianred3", vertex.size=5, vertex.frame.color="ivory1", 
                vertex.label.cex=0.8, vertex.label.dist=5, edge.curved=0.2)
  
  # Plot with community detection
  if(community) {
    cfg = cluster_fast_greedy(G)
    params = append(list(cfg), params)
  }
  do.call(plot, params)
}

plot_comparison_graphs = function(G_bonferroni, G_raw, community=F) {
  # Plotting graphs
  par(mfrow = c(2,2), mai=c(.1,.1,.2,.1))
  graph_list = list(G_bonferroni$A, G_bonferroni$B, G_raw$A, G_raw$B)
  title_list = c("Group A (Bonferroni)", "Group B (Bonferroni)",
                 "Group A (non-adjusted)", "Group B (non-adjusted)")
  invisible(sapply(1:4, function(i) plot_graph(graph_list[[i]], title_list[[i]],community=community)))
}
```


## Define che correlation matrices 
In this section we preprocess the data we need in order to apply the methods defined above. Firstly we create the two correlation matrices using the Pearson correlation, then to simplify operations we 'flat' the matrices and we pass them to the `build_graph` to obtain the two types: adjusted and not adjusted.
```{r test, warning=FALSE}
# correlation matrices
A = lapply(asd_sel, function(x) cor(x, method = "pearson"))
B = lapply(td_sel, function(x) cor(x, method = "pearson"))

# flatted correlation matrices
A_flat = t(sapply(A, unlist))
B_flat = t(sapply(B, unlist))

# building adjacency matrix for ASD group
G_bonferroni = build_graph(A_flat, B_flat)
G_raw = build_graph(A_flat, B_flat, adjust = F)
```

## Visualize the estimated graphs
```{r graphs, echo=FALSE, warning=FALSE}
plot_comparison_graphs(G_bonferroni, G_raw)
plot_comparison_graphs(G_bonferroni, G_raw, community = T)
```

## Paired correlation between groups
Until now we tested the correlations within groups with a specific threshold, in the following we look at the number of concomitant correlations above threshold between groups. To do so we first build the individual graphs and then compare their adjacency matrix, a plot of the number of correlations the two groups agree on (i.e. both above threshold) is then plotted.
```{r th, warning=FALSE}
# Paired correlation with sliding threshold
probs = seq(0.5, 1, 0.05)
n = length(probs)
n_coact_bonf = rep(NA, n)
n_coact_raw = rep(NA, n)

for(i in 1:n) {
  G_bonf = build_graph(A_flat, B_flat, probs = probs[i], adjust = T)
  G_raw = build_graph(A_flat, B_flat, probs = probs[i], adjust = F)
  
  # number of paired correlation between groups
  n_coact_bonf[i] = sum((G_bonf$A == 1) & (G_bonf$B == 1)) - 116
  n_coact_raw[i] = sum((G_raw$A == 1) & (G_raw$B == 1)) - 116
}
```


```{r graph, echo=FALSE}
par(mfrow = c(1,1), mai=c(1,1,1,1))
plot(x = probs, y = n_coact_bonf, type = "b", col = "seagreen3", pch = 16, xlab = "Threshold",
     ylab = "Number of Paired Correlations", ylim = range(n_coact_raw), lwd = 2,
     main = "Paired correlations at different thresholds")
lines(x = probs, y = n_coact_raw, type = "b", col = "indianred3", pch = 16, lwd = 2)
segments(.5, n_coact_bonf[7], 1, n_coact_bonf[7], lwd = 2, lty = 2, col = "lightcyan4")
text(.8, n_coact_bonf[7] - 150, bquote(q[80]),lwd = 2)
legend("topright", legend=c("Bonferroni", "Non-Adjusted"),
       col=c("seagreen3", "indianred3"), lty=1, lwd=3)
grid()
```

As we can notice from the plot, when the threshold goes up, the number of edges goes down. That’s reasonable since using higher value of threshold means accepting only a stronger association between the features. To emphasize this, we also show the degree distributions of the difference graph: we can see how it changes even if the the smallest and largest thresholds differ by only 0.3.

```{r boot_plot, warning=FALSE}
# Difference graph
A_means = colMeans(A_flat)
B_means = colMeans(B_flat)
G_delta = abs(A_means - B_means)

# Plot of differences at different thresholds
t_range = quantile(G_delta, probs = c(0.85, 0.95))
t_diff = seq(t_range[1], t_range[2], length.out = 4)
adj_matrix = function(t) as.integer(G_delta >= t)
delta_matrices = t(sapply(t_diff, adj_matrix))
titles = unlist(lapply(t_diff, function(t) paste(c("t =", round(t,2)), collapse = " ")))

plot_degree_distributions = function(flat_matrices, titles) {
  colors = c("deeppink", "deepskyblue", "gold2", "orchid")
  for(i in 1:4) {
    M = to_matrix(flat_matrices[i,])
    G = graph_from_adjacency_matrix(M, mode="undirected")
    deg = degree(G, mode="all")
    deg.dist = degree.distribution(G, cumulative=T, mode="all")
    if(i == 1) {
      plot(x=0:max(deg), y=1-deg.dist, pch=19, type="l", col=colors[i],
           xlab="Degree", ylab="Cumulative frequency",
           main="Degree distribution at different thresholds" , lwd=2)
    }
    else {
      lines(x=0:max(deg), y=1-deg.dist,
            col=colors[i], lwd=2)
    }
  }
  grid()
  legend("bottomright", legend=titles, col=colors, lty=1, lwd=2)
}

par(mfrow=c(1,1), mai=c(1,1,1,1))
plot_degree_distributions(delta_matrices, titles)
```


## Conclusions
Through the difference graph we can visualize clearly the co-activation differences between the two groups. The graph has been built taking the adjacency matrix of both groups placing an edge where they differ, i.e. a pair of features is strongly correlated in one group and not in the other. What we can see from this graph is that:

- there are a couple of large groups of features on which the two group disagree.
- in particular this larger cmomunities also show a different behaviour with many connections between features while the rest look more like paths
- there are plenty of singleton features meaning that any potential (un)correlation between those is seen in both groups

```{r}
G_diff = abs(G_bonferroni$A - G_bonferroni$B)
par(mfrow=c(1,1), mai=c(.5,.5,.5,.5))
plot_graph(G_diff, title = "Difference Graph", community = T)
```

## Bootstrap

In order to build bootstrap confidence intervals for the correlation difference we sample multiple times subjects in both groups, compute the average correlation and keep track of the difference in absolute value for each pair of features. Finally we build a pivotal percentile interval, we slide thresholds for different values noticing values around the first quartile allowed to retain enough edges to see patterns in the difference graph.
Both normal and percentile have been tried but the resulting CI were too large hence intersecting very often even with low thresholds.
It is difficult to draw a definitive conclusion from these graphs but considering the max difference is around 0.35, it seems that even a tiny threshold like 0.1 doesn't capture much of a difference in the correlation between features from the two groups.
While neuroscience is an area obscure to us we expected to see a more marked difference in the two groups. 

```{r}
b = 300

# point estimate correlation
cor_a = colMeans(A_flat)
cor_b = colMeans(B_flat)

diff = abs(cor_a - cor_b)

t = quantile(diff, probs = .3, na.rm = T)

D_boot = matrix(NA, nrow = b, ncol = 116^2)
n = 12

d = 116
m = d*(d-1)/2
alpha = 0.05/m

for(i in 1:b) {
  idx_a = sample(1:n, n, replace = T)
  idx_b = sample(1:n, n, replace = T)
  
  D_boot[i,] = abs(colMeans(A_flat[idx_a,]) - colMeans(B_flat[idx_b,]))
}

# Percentile interval
ci_perc = matrix(NA, nrow=2, ncol=116^2)
ci_perc[1,] = 2*diff - apply(D_boot, MARGIN = 2, function(x) quantile(x, probs = 1-alpha))
ci_perc[2,] = 2*diff - apply(D_boot, MARGIN = 2, function(x) quantile(x, probs = alpha))

# Adjacency matrices
t = quantile(diff, probs=c(0.2,.4,.6,.75), na.rm=T)
par(mfrow=c(2,2), mai=c(.1,.1,.3,.1))

for(i in 1:4) {
  M_diff = get_adjancency_matrix(ci_perc[1,], ci_perc[2,], t[i])
  plot_graph(M_diff, title = paste("t = ", round(t[i], 2)))
}

```

